Supplemental Material for: The evolution of gene expression in the term placenta of viviparous mammals

### 2016-09-28 manuscript/manuscript.Rnw
add core placenta plot to manuscript

%\documentclass[english,12pt]{article}
\documentclass[5p]{elsarticle_dla}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\setmathrm{FreeSerif}
\setmathsf{FreeSans}
\setmathtt{FreeSerif}
%\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage[letterpaper]{geometry}
%\usepackage{url}
\usepackage{graphicx}
%\usepackage{rotating}
%\usepackage{wrapfig}
\usepackage{booktabs}
%\usepackage{float}
\usepackage{multirow}
%\usepackage{setspace}
\usepackage{acronym}
\usepackage{array}
\usepackage{longtable}
\usepackage{amsmath}
\usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black,citecolor=blue]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads]{endfloat}
% \renewcommand{\efloatseparator}{\mbox{}}
%\AtBeginFigures{\newpage \renewcommand{\efloatseparator}{\mbox{}}\section*{Figure Legends}}
%\AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
% \renewcommand{\includegraphics}[2][]{}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\newcommand{\DEW}[1]{\textcolor{purple}{\fxnote{DEW: #1}}}
% \usepackage[x11names,svgnames]{xcolor}
% \usepackage{texshade}
% \usepackage{tikz}
\usepackage{nameref}
\usepackage{xspace}
% this package will stretch the font slightly if necessary.
% \usepackage{microtype}
\usepackage{zref-xr,zref-user}
\zxrsetup{toltxlabel}
\zexternaldocument*{supplemental_material}
\usepackage[capitalize]{cleveref}

@@

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all species studied and was
not a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
\cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>1 & 
`Minimum Expression in` == '14 of 14 species',max(num_genes)]} %
genes in the core transcriptome.
%
Multiple components of annexin complexes are included in the core
transcriptome, including \textit{ANXA2}, \textit{ANXA1},
\textit{S100A11}, \textit{S100A10}. ANXA1 is involved in the
resolution of the inflammation response through inhibition of
phospholipase A2 activity and signaling through the formyl peptide
receptor family, and has previously been hypothesized to be important
for the maintenance of the anti-inflammatory state during pregnancy
\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
known to interact with S100A10, S100A6, S100A11, and S100A4
\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
its interaction with S100P as well; it is involved in cell-cell
interaction as well as vesicle trafficking and von Willibrand Factor
secretion. The expression regulation of another core gene,
\textit{TIMP3}, has been implicated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
Supplemental \cref{tab:core_placenta_go_terms} lists the significant
enrichment of \ac{GO} terms in the core placenta transcriptome;
notable terms include those involved in cytoskeletal structure,
protein transport, membrane protein structure, and inflammation
resolution. Increasing the permissiveness of the core placenta
transcriptome by only requiring 10 of 14 species to express 1:1
orthologs with FPKM $>10$ and results in the same terms also being
significantly enriched, along with additional terms (Supplemental
\cref{tab:core_placenta_go_terms}).

>>>>

%\documentclass[english,12pt]{article}
\documentclass[5p]{elsarticle_dla}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\setmathrm{FreeSerif}
\setmathsf{FreeSans}
\setmathtt{FreeSerif}
%\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
%\usepackage[letterpaper]{geometry}
%\usepackage{url}
\usepackage{graphicx}
%\usepackage{rotating}
%\usepackage{wrapfig}
\usepackage{booktabs}
%\usepackage{float}
\usepackage{multirow}
%\usepackage{setspace}
\usepackage{acronym}
\usepackage{array}
\usepackage{longtable}
\usepackage{amsmath}
\usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black,citecolor=blue]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads]{endfloat}
% \renewcommand{\efloatseparator}{\mbox{}}
%\AtBeginFigures{\newpage \renewcommand{\efloatseparator}{\mbox{}}\section*{Figure Legends}}
%\AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
% \renewcommand{\includegraphics}[2][]{}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\newcommand{\DEW}[1]{\textcolor{purple}{\fxnote{DEW: #1}}}
% \usepackage[x11names,svgnames]{xcolor}
% \usepackage{texshade}
% \usepackage{tikz}
\usepackage{nameref}
\usepackage{xspace}
% this package will stretch the font slightly if necessary.
% \usepackage{microtype}
\usepackage{zref-xr,zref-user}
\zxrsetup{toltxlabel}
\zexternaldocument*{supplemental_material}
\usepackage[capitalize]{cleveref}

@@

<<core.placenta.transcriptome.distribution,results="hide",echo=FALSE,message=FALSE>>=
num.species <- star.logs.wide[,length(unique(species))]
calculate.distribution.shape <- 
    function(min.vector) {
        min.vector <- sort(min.vector,decreasing=TRUE)
        temp <- data.table(FPKM=min.vector,num_genes=1:length(min.vector))
        temp <- temp[order(-num_genes)][!duplicated(FPKM)]
        return(temp)
    }

cptg.nh <- core.placenta.transcriptome.genes.shape[!(human_name %in% housekeeping.genes.superset[,gene_short_name])]

cpt.distribution.shape <- 
    lapply(c("all_expression",
             "all_but_one_expression",
             "all_but_two_expression",
             "all_but_three_expression",
             "all_but_four_expression"),
           function(x){
               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Minimum Expression in`:=x]
           })
fix.expression.names <- function(x){
    x <- gsub("^all_expression$",
              paste0(num.species," of ",num.species," species"),
              x
              )
    nc.levels <- paste0(num.species," of ",num.species," species")
    ## number coutner
    nc <- c("one"=1,"two"=2,"three"=3,"four"=4)
    for (a in names(nc)) {
        x <- gsub(paste0("^all_but_",a,"_expression$"),
              paste0(num.species-nc[a]," of ",num.species," species"),
              x
              )
        nc.levels <-
            c(nc.levels,
              paste0(num.species-nc[a]," of ",num.species," species"))
    }
    return(factor(x,levels=nc.levels))
}

cpt.distribution.shape <- 
    rbindlist(cpt.distribution.shape)

cpt.distribution.shape <- 
    cpt.distribution.shape[,`Minimum Expression in`:=
                                fix.expression.names(`Minimum Expression in`)]

setnames(cpt.distribution.shape,
         "FPKM",
         "Expression (FPKM)")
@

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all species studied and was
not a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
Supplemental \cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>1 & 
`Minimum Expression in` == '14 of 14 species',max(num_genes)]} %
genes in the core transcriptome.
%
Multiple components of annexin complexes are included in the core
transcriptome, including \textit{ANXA2}, \textit{ANXA1},
\textit{S100A11}, \textit{S100A10}. ANXA1 is involved in the
resolution of the inflammation response through inhibition of
phospholipase A2 activity and signaling through the formyl peptide
receptor family, and has previously been hypothesized to be important
for the maintenance of the anti-inflammatory state during pregnancy
\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
known to interact with S100A10, S100A6, S100A11, and S100A4
\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
its interaction with S100P as well; it is involved in cell-cell
interaction as well as vesicle trafficking and von Willibrand Factor
secretion. The expression regulation of another core gene,
\textit{TIMP3}, has been implicated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
Supplemental \cref{tab:core_placenta_go_terms} lists the significant
enrichment of \ac{GO} terms in the core placenta transcriptome;
notable terms include those involved in cytoskeletal structure,
protein transport, membrane protein structure, and inflammation
resolution. Increasing the permissiveness of the core placenta
transcriptome by only requiring 10 of 14 species to express 1:1
orthologs with FPKM $>10$ and results in the same terms also being
significantly enriched, along with additional terms (Supplemental
\cref{tab:core_placenta_go_terms}).

### 2016-09-28 manuscript/supplemental_material.Rnw
add core placenta plot to manuscript

<<placenta_class_table,echo=FALSE,results="hide">>=
placenta_class_table <- function(class="shape") {
    subtable <- placenta.classification.p[factor==class & type == "aov",list(gene,p,fdr)]
    setnames(subtable,c("Gene","p","FDR"))
    subtable <- head(subtable[FDR <= 0.05,][order(p),],n=50)
    print(xtable(subtable,
                 align=c("c","c","r","r"),
                 display=c("s","s","g","g")
                 ),
          booktabs=TRUE,
          floating=FALSE,
          include.rownames=FALSE,
          math.style.exponents=TRUE,
          size="\\tiny")
}

@@

\begin{table}
<<attachment_shape_significant,echo=FALSE,results="asis">>=
placenta_class_table("shape")
@ 
\caption{Top 50 genes which are associated with the shape of placenta attachment}
\label{tab:shape_genes}
\end{table}

@@

<<intimacy_significant,echo=FALSE,results="asis">>=
placenta_class_table("barrier")
@ 
\caption{Top 50 genes which are associated with the degree of placenta intimacy}
\label{tab:barrier_genes}
\end{table}

@@

<<interdigitation_significant,echo=FALSE,results="asis">>=
placenta_class_table("interdigitation")
@ 
\caption{Top 50 genes which are associated with the degree of maternal-fetal interdigitation}
\label{tab:interdigitation_genes}
\end{table}

@@

\subsection{Pathway Analysis}
\begin{table}
<<pathway_analysis,echo=FALSE,results="asis">>=
pathway <- list()
for (database in c("kegg","reactome","biocarta","netpath")) {
    load(paste0("../data/placenta_gsea_",database))
    eval(parse(text=paste0("gsea.table <- placenta_gsea_",database)))
    gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
    pathway[[database]] <-
        gsea.table
}
pathway.analysis <-
    rbindlist(pathway)[fdr < 0.1,]
rm(pathway)
pathway.analysis[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
                                gsub("_+"," ",
                                     Pathway))]
pathway.analysis[,Pathway:=gsub(".+\\%","",Pathway)]

setnames(pathway.analysis,"p_values","p")
setorder(pathway.analysis,"p")
print(xtable(pathway.analysis,
             display=c("s","s","f","f","s"),
             align=c("c","p{2in}","r","r","c")),
      math.style.exponents=TRUE,
      include.rownames=FALSE,
      booktabs=TRUE,
      floating=FALSE)      
@ 
\caption[Core Placenta Pathway Enrichment]{Pathways with significant and suggestive enrichment
  ($\text{FDR}\le 0.1$) in the core placenta transcriptome using the
  biocarta, reactome, and kegg pathway databases}
\label{tab:core_placenta_pathway_analysis}
\end{table}

\begin{table}
<<pathway_analysis_class,echo=FALSE,results="asis">>=
pathway <- list()
for (class in c("shape","interdigitation","barrier")) {
    for (database in c("kegg","reactome","biocarta","netpath")) {
        load(paste0("../data/placenta_gsea_",class,"_",database))
        eval(parse(text=paste0("gsea.table <- placenta_gsea_",class,"_",database)))
        gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
        gsea.table[,Class:=gsub("^(.)","\\U\\1",class,perl=TRUE)]
        pathway[[paste0(class,"_",database)]] <-
            gsea.table
    }
}
placenta_class_gsea <-
    rbindlist(pathway)[fdr < 0.1,]
placenta_class_gsea[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
                                   gsub("_+"," ",
                                        Pathway))]
placenta_class_gsea[,Pathway:=gsub(".+\\%","",Pathway)]
setnames(placenta_class_gsea,"p_values","p")
setorder(placenta_class_gsea,"p")
print(xtable(placenta_class_gsea,
             display=c("s","s","f","f","s"),
             align=c("c","p{2in}","r","r","c")),
      math.style.exponents=TRUE,
      include.rownames=FALSE,
      booktabs=TRUE,
      floating=FALSE)      
@ 
\caption[Placenta Classification Pathway Enrichment]{Pathways with
  significant and suggestive enrichment ($\text{FDR}\le 0.1$) in genes
  associated with placenta morphology using the biocarta, reactome,
  and kegg pathway databases.}
\label{tab:core_placenta_pathway_analysis}
\end{table}

>>>>

<<placenta_class_table,echo=FALSE,results="hide">>=
placenta_class_table <- function(class="shape") {
    subtable <- placenta.classification.p[factor==class & type == "aov",list(gene,p,fdr)]
    setnames(subtable,c("Gene","p","FDR"))
    subtable <- head(subtable[FDR <= 0.05,][order(p),],n=50)
    print(xtable(subtable,
                 align=c("c","c","r","r"),
                 display=c("s","s","g","g")
                 ),
          booktabs=TRUE,
          floating=FALSE,
          include.rownames=FALSE,
          math.style.exponents=TRUE,
          size="\\scriptsize")
}

@@

\begin{table}[h]
<<attachment_shape_significant,echo=FALSE,results="asis">>=
placenta_class_table("shape")
@ 
\caption{Top 50 genes which are associated with the shape of placenta attachment}
\label{tab:shape_genes}
\end{table}
\clearpage

@@

\begin{table}[h]
<<intimacy_significant,echo=FALSE,results="asis">>=
placenta_class_table("barrier")
@ 
\caption{Top 50 genes which are associated with the degree of placenta intimacy}
\label{tab:barrier_genes}
\end{table}

@@

\begin{table}[h]
<<interdigitation_significant,echo=FALSE,results="asis">>=
placenta_class_table("interdigitation")
@ 
\caption{Top 50 genes which are associated with the degree of maternal-fetal interdigitation}
\label{tab:interdigitation_genes}
\end{table}

@@

\subsection{Pathway Analysis}
\begin{table}
<<pathway_analysis,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
pathway <- list()
for (database in c("kegg","reactome","biocarta","netpath")) {
    load(paste0("../data/placenta_gsea_",database))
    eval(parse(text=paste0("gsea.table <- placenta_gsea_",database)))
    gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
    pathway[[database]] <-
        gsea.table
}
pathway.analysis <-
    rbindlist(pathway)[fdr < 0.1,]
rm(pathway)
pathway.analysis[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
                                gsub("_+"," ",
                                     Pathway))]
pathway.analysis[,Pathway:=gsub(".+\\%","",Pathway)]

setnames(pathway.analysis,"p_values","p")
setorder(pathway.analysis,"p")
print(xtable(pathway.analysis,
             display=c("s","s","g","g","s"),
             align=c("c","p{3.5in}","r","r","c")),
      math.style.exponents=TRUE,
      include.rownames=FALSE,
      booktabs=TRUE,
      floating=FALSE)      
@ 
\caption[Core Placenta Pathway Enrichment]{Pathways with significant and suggestive enrichment
  ($\text{FDR}\le 0.1$) in the core placenta transcriptome using the
  biocarta, reactome, and kegg pathway databases}
\label{tab:core_placenta_pathway_analysis}
\end{table}

\begin{table}
<<pathway_analysis_class,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
pathway <- list()
for (class in c("shape","interdigitation","barrier")) {
    for (database in c("kegg","reactome","biocarta","netpath")) {
        load(paste0("../data/placenta_gsea_",class,"_",database))
        eval(parse(text=paste0("gsea.table <- placenta_gsea_",class,"_",database)))
        gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
        gsea.table[,Class:=gsub("^(.)","\\U\\1",class,perl=TRUE)]
        pathway[[paste0(class,"_",database)]] <-
            gsea.table
    }
}
placenta_class_gsea <-
    rbindlist(pathway)[fdr < 0.1,]
placenta_class_gsea[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
                                   gsub("_+"," ",
                                        Pathway))]
placenta_class_gsea[,Pathway:=gsub(".+\\%","",Pathway)]
setnames(placenta_class_gsea,"p_values","p")
setorder(placenta_class_gsea,"p")
print(xtable(placenta_class_gsea,
             display=c("s","s","g","g","s","s"),
             align=c("c","p{3in}","r","r","c","c")),
      math.style.exponents=TRUE,
      include.rownames=FALSE,
      booktabs=TRUE,
      floating=FALSE)      
@ 
\caption[Placenta Classification Pathway Enrichment]{Pathways with
  significant and suggestive enrichment ($\text{FDR}\le 0.1$) in genes
  associated with placenta morphology using the biocarta, reactome,
  and kegg pathway databases.}
\label{tab:core_placenta_pathway_analysis}
\end{table}
